Effects of pulsed hollow electron lens operation on the beam core in HL-LHC: \\First experimental studies and simulations.

### 2018-01-29 resex.tex
remove grid lines

\begin{abstract}
In the HL-LHC a considerable amount of energy is stored in the beam tails due to the high beam intensity and an overpopulation of the tails compared to a Gaussian distribution. To control and clean the tail population, the installation of two hollow electron lenses, one per beam, is considered. Beside the DC operation, also a pulsed operation of the hollow electron lens is considered, which would considerably increase the diffusion speed by putting noise on the halo particles. In the ideal case, that is in case of no field at the beam core, only the halo particles are excited while leaving the core unperturbed. The picture though changes, if a residual field is present also at the location of the beam core putting noise also on the beam core. In this paper we present for estimates of the residual field at the beam core expected from the HL-LHC hollow electron lens and first experimental results of the effect of this excitation on the beam core together with the supporting simulations.
\end{abstract}

@@

All of the above reasons lead to the conclusion that a mechanism is needed to deplete the beam tails in a controlled manner (see for example \cite{helreview} in case of HL-LHC). The most obvious idea is to decrease the collimator gaps or scrape the tails with a collimator type device. The minimum gap size of the collimators is however limited by the impedance of the machine and small enough gap sizes can therefore not be reached without risking an instability of the beam. Other mechanisms must therefore be found to actively deplete the tails.

Most promising are methods, which considerably increase the diffusion speed in the region of the halo particles while leaving the beam core unperturbed. The diffusing halo particles are then intercepted by the collimation system and removed (see Fig.~\ref{fig:active_halo_control} for illustration). This concept is also referred to as active halo control compared to a passive system, which only intercepts the halo particles without actively controlling the diffusion speed. Currently, the LHC collimation system is a passive halo control system. 
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{passive_halo_control.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{active_halo_control.png}
	\end{minipage}	
	\caption{\label{fig:active_halo_control} Sketch of passive halo control as with the collimation system (top) and active halo control using in addition for example a hollow electron lens (HEL) to control the diffusion speed in the region tail region without affected the beam core (bottom).}
\end{figure}

In view of the need of an active halo control system for LHC and HL-LHC in particular and in general for future high power accelerators, different methods have been studied in the last recent years \cite{helreview_bruce}, of which the HEL is considered the superior device at least in case of the HL-LHC \cite{helreview} and is also considered for other future high energy colliders like HE-LHC and FCC-hh \cite{helhcparam2011,fcc_coll_ipac2017}.

@@

This paper is structured as follows: Section~\ref{sec:hel} gives an introduction to the concept of HELs and summarizes the design parameters of the HL-LHC HEL. Section~\ref{sec:core} is dedicated to describing the sources of a residual field from the HEL in the core region. Section~\ref{sec:exp_sum}--\ref{sec:exp_sim} describes then in detail the two beam experiments conducted at the LHC to study the effect on the beam core in case of a pulsed HEL operation. This includes the detailed analysis of the observed losses, emittance and transverse beam distribution changes. To the knowledge of the authors, the observed distribution changes presented in this paper have never been measured before in the context of a resonant excitation. A resonant excitation has been previously studied at the Tevatron in the context of the HEL experiments \cite{hel_tevatron_stancari} and the abort gap cleaning used in operation \cite{hel_tevatron_abortgap_zhang}. Both studies however only concentrated on the losses and emittance changes without measuring the detailed changes of the distribution itself. In addition, the presented experiments also provide scalings of the losses and emittance growth with the excitation amplitude. This allows to provide first tolerances for the residual field of the HL-LHC HEL at the beam core and also to compare the obtained results with simulations and thus benchmark the same.

>>>>

\begin{abstract}
In case of the High Luminosity Large Hadron Collider (HL-LHC), a considerable amount of energy is stored in the beam tails due to the high beam intensity and, in addition, an overpopulation of the tails compared to a Gaussian distribution. To control and clean the tail population, the installation of two hollow electron lenses, one per beam, is foreseen. Beside the DC operation of the lenses, also pulsed operation is considered having the advantage to considerably increase the diffusion speed by exerting noise on the halo particles. In the ideal case, that is in case of no field at the proton beam core, only the halo particles are excited while leaving the core unperturbed. The picture though changes, if a residual field is present also at the location of the beam core exposing also the the particles in the center to noise. In this paper, we present first estimates of the residual field at the proton beam core expected from the HL-LHC hollow electron lens and summarize the results of the LHC experiments conducted in 2016 and 2017 to study the effects on the beam core in case of pulsed operation.
\end{abstract}

@@

All of the above reasons lead to the conclusion that a mechanism is needed to deplete the beam tails in a controlled manner \cite{helreview}. The most obvious idea is to decrease the collimator gaps or scrape the tails with a collimator type device. The minimum gap size of the collimators is however limited by the impedance of the machine. Small enough gap sizes can therefore not be reached without risking severe beam instabilities. Other mechanisms must therefore be found to actively deplete the tails. Most promising are methods, which increase the diffusion speed in the region of the halo particles while leaving the beam core unperturbed. The diffusing halo particles are then intercepted by the collimation system and removed. This concept is also referred to as active halo control compared to a passive system, which only intercepts the halo particles without actively controlling the diffusion speed (see Fig.~\ref{fig:active_halo_control} for illustration).
\begin{figure}[h]
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		\includegraphics[width=0.7\linewidth]{passive_halo_control.png}
	\end{minipage}
	\begin{minipage}[t]{1.0\linewidth}
		\centering
		\includegraphics[width=0.7\linewidth]{active_halo_control.png}
	\end{minipage}	
	\caption{\label{fig:active_halo_control} Sketch of passive halo control as with the collimation system (top) and active halo control using in addition for example a hollow electron lens (HEL) to control the diffusion speed in the region tail region without affected the beam core (bottom).}
\end{figure}

In a recent review, the need for such an active halo control system for HL-LHC has been assessed with the conclusion that it would considerably increase the margin and reduce the risk for machine protection compared to the current LHC collimation system being a passive halo control system  \cite{helreview}. In view of the general need of active halo control for future high power accelerators and HL-LHC in particular, different methods have been studied in the last recent years \cite{helreview_bruce}, of which the HEL is considered the superior device at least in case of the HL-LHC \cite{helreview} and is also considered for other future high energy colliders like HE-LHC and FCC-hh \cite{helhcparam2011,fcc_coll_ipac2017}.

@@

This paper is structured as follows: Section~\ref{sec:hel} gives an introduction to the concept of HELs and summarizes the design parameters of the HL-LHC HEL. Section~\ref{sec:core} is dedicated to describing the sources of a residual field from the HEL in the core region. Section~\ref{sec:exp} describes then in detail the two beam experiments conducted at the LHC to study the effect on the beam core in case of a pulsed HEL operation. This includes the detailed analysis of the observed losses, emittance and transverse beam distribution changes. To the knowledge of the authors, the observed distribution changes presented in this paper have never been measured before in the context of a resonant excitation. A resonant excitation has been previously studied at the Tevatron in the context of the HEL experiments \cite{hel_tevatron_stancari} and the abort gap cleaning used in operation \cite{hel_tevatron_abortgap_zhang}. Both studies however only concentrated on the losses and emittance changes without measuring the detailed changes of the distribution itself. In addition, the presented experiments also provide scalings of the losses and emittance growth with the excitation amplitude. This allows to provide first tolerances for the residual field of the HL-LHC HEL at the beam core and also to compare the obtained results with simulations and thus benchmark the same.
